---
title: "Visualisation of Household Income Data of Singapore"
author: "YuXuan Tay"
date: "Sunday, February 08, 2015"
output:
  html_document:
    fig_height: 6
    fig_width: 8
    keep_md: yes
    toc: yes
---

```{r options, echo=FALSE}
knitr::opts_chunk$set(results = "hold", fig.show = "hold", message = F, warning = F)
```

## Introduction

In this exercise, I will attempt to create a few visualisation of the household income data of Singapore.
The purpose of this exercise is to gain an understanding of the income distribution and trend in Singapore.
Plots will be drawn with the `ggvis` visualisation package, which produces SVG graphics.
This means they will be scalable to the reader's discretion.

The data is obtained from [data.gov.sg](http://data.gov.sg/), 
which is the national database for public government data. 
Specifically, I will be exploring the average household income and distribution data prepared by the
Singapore Department of Statistics ([SingStats](http://www.singstat.gov.sg/)). The data contained information in the years 2000 to 2013.

## Data

```{r libraries}
# load libraries
library(tidyr); library(dplyr); library(stringr)
library(ggplot2); library(ggvis)
library(knitr)
```

The 3 main datasets I will be looking at are as follows:

- average and median monthly household income
- average monthly household income by deciles
- households by monthly household income

SingStats prepared a few similar datasets depending on the definitions of households and income.
For households, the following definitions are available:

- resident households
- resident **employed** households
- resident households **per household member**
- resident **employed** households **per household member**

For income,

- income **including** employer CPF contributions
- income **excluding** employer CPF contributions

For consistency, I have used the datasets for resident employed households 
and for income excluding employer CPF contributions. 
That is because I am interested in the income distribution of Singapore and would like to exclude the unemloyed.
Excluding employer CPF contributions would be to do a fair comparison since foreigners are also included
and they do not receive these contributions.

## Average and Median Monthly Household Income

Let's first look at the average and median monthly household income.
All the data presented in this page is obtained directly from the [data.gov.sg](https://data.gov.sg) website
with minimal processing to format the column names and save it into tab-separated text files.
Note that the income amounts presented are nominal numbers without accounting for inflation.

```{r average.data, results = 'asis'}
income.average <- read.table('average_and_median_monthly_household_income.txt', header = T)
kable(income.average, caption = 'Table of Mean and Median Household Income')
```

### Mean and Median Income Levels

Let's look at how this appears in a graph.

```{r average.line, results = 'asis'}
income.average %>%
    select(year, mean, median) %>%
    ggvis(~factor(year)) %>%
    layer_lines(y = ~mean, stroke = "Mean") %>%
    layer_lines(y = ~median, stroke = "Median") %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Income', title_offset = 60) %>%
    add_legend('stroke', title = 'Measure')
```

A few observations can be made. From year 2000 to 2004, the average income levels were quite flat. 
This was possibly due to the dot-com bust, which led to a recession in Singapore. 
From 2004 onwards, income levels increased quickly except during 2009 due to the subprime mortgage crisis.

Another observation is that the mean income is substantially higher than the median income.
This is one of the arguments why mean income is not a good indicator of income levels in a country.
Mean income is not a robust measure of average as it is affected by extreme values.
A right-skewed income distribution will shift the mean income level even though 
this does not mean an elevated income for the majority of the population.
The skewed income distribution of Singapore will be more evident later when we explore the average income by deciles.
In general, it is always a good idea to be wary of mean or average values if the distribution of 
the variable is not clearly known.

On the other hand, the median income level gives the amount that separates the population of
income earners into half: 50% of them earn less than the median level and the other 50% earn more.
It is robust to extreme values since changes at both ends of the distribution do not affect it.

```{r average.index}
income.average <- mutate(income.average,
                         mean.index.2000 = mean / first(mean),
                         median.index.2000 = median / first(median))
```

The nominal mean and median income levels have grown by 
`r round((last(income.average$mean.index.2000) - 1) * 100, digits = 2)`% and
`r round((last(income.average$median.index.2000) - 1) * 100, digits = 2)`% respectively
in the 13-year period, with a compounded annual growth of
`r round((last(income.average$mean.index.2000) ^ (1/13) - 1) * 100, digits = 2)`% and
`r round((last(income.average$median.index.2000) ^ (1/13) - 1) * 100, digits = 2)`% respectively.

### Ratio of Mean to Median Income Levels

Let us now examine the ratio of mean to median levels across the years.

```{r mean.median.ratio, results = 'asis'}
# compute mean-median ratio
income.average <- mutate(income.average, mean.median.ratio = mean / median)
income.average %>%
    select(year, mean.median.ratio) %>%
    ggvis(~factor(year), ~mean.median.ratio) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Mean-Median Ratio', title_offset = 50)
```

At first glance, it seems that, except for 2013, the mean-median ratio has been in a general increasing trend, 
which means that high earners are growing their income faster than low earners. 
However, the increase has been relatively low, at
`r with(income.average, round((max(mean.median.ratio) / min(mean.median.ratio) - 1) * 100, digits = 2))`%
over the 12-year period. This suggests that the income distribution might be quite consistent over the period.
I think including zero into the y-axis will make this more obvious, but have not figured out how to 
set the limits of the axes with `ggvis`. 
This is also the reason why the axes in all other plots are unadjusted even though 
it would be a good idea to include zero in the y-axis in a few other plots to give a right sense of scale.

## Average Monthly Household Income Distribution by Deciles

A better way to explore the income distribution is to look at the average monthly income by deciles.
Do note that in this data, the numbers are for the average income by deciles and **not** of income at decile levels.
In other words, the numbers for decile 1 is for the average income for the bottom 10 percentile of households,
and not the 10th percentile level of income.
The 10th percentile level of income would actually represent the maximum income of 
the bottom 10 percentile of household.

```{r decile.data, results = 'asis'}
income.decile <- read.table('average_monthly_household_income_by_deciles.txt', header = T)
kable(income.decile, caption = 'Table of Average Household Income by Decile')
```

I would first do some transformation to the data to compute a few other variables of interest.
It would also be necessary for me to reshape it from wide to long format with the `tidyr` package.

```{r decile.transform}
# convert to long format
income.decile <- gather(income.decile, decile, income, -year)
# convert decile from string into numeric values
income.decile <- mutate(income.decile,
                        decile = as.numeric(str_extract(decile, '\\d+')))
# merge average dataset with decile dataset
income.average <- income.average %>%
                    select(year, mean, median) %>%
                    gather(decile, income, -year)
income.decile <- rbind(income.decile, income.average)
# format factor values
income.decile <- mutate(income.decile, decile = factor(decile, levels = c(1:10, 'mean', 'median')))
```

### Line Graph of Average Income by Deciles

This is how the original data looks like in a graph. 
The mean and median income levels have been included for comparisons.

```{r decile.line, results = 'asis'}
income.decile %>%
    select(year, decile, income) %>%
    mutate(strokeWidth = ifelse(decile %in% c('mean', 'median'), 2, 1),
           opacity = ifelse(decile %in% c('mean', 'median'), 0.7, 1)) %>%
    ggvis(~factor(year), ~income, stroke = ~factor(decile), 
          strokeWidth := ~strokeWidth, opacity := ~opacity) %>%
    group_by(decile) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Average Income', title_offset = 60) %>%
    add_legend(c('stroke', 'strokeWidth','opacity'), title = 'Decile')
```

It immediately becomes obvious how skewed the income distribution of Singapore is,
with the top decile earners making substantially more than even the 2nd highest decile earners.
The general trend across the deciles is similar to the overall average income level.

Curiously, the top decile suffered a significant income drop in year 2013, 
while the other deciles were unaffected. I am puzzled as to the exact cause.

### Income Growth from Year 2000 by Deciles

For better comparision of growth, I will normalise the income levels of each decile to the year 2000 levels.

```{r decile.index, results = 'asis'}
# normalise average income of each decile to year 200 levels
income.decile <- income.decile %>%
                    group_by(decile) %>%
                    arrange(year) %>%
                    mutate(index.2000 = income / first(income)) %>%
                    ungroup()
income.decile %>%
    select(year, decile, index.2000) %>%
    mutate(strokeWidth = ifelse(decile %in% c('mean', 'median'), 2, 1),
           opacity = ifelse(decile %in% c('mean', 'median'), 0.7, 1)) %>%
    ggvis(~factor(year), ~index.2000, stroke = ~factor(decile), 
          strokeWidth := ~strokeWidth, opacity := ~opacity) %>%
    group_by(decile) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Income Index to Year 2000 Levels', title_offset = 40) %>%
    add_legend(c('stroke', 'strokeWidth','opacity'), title = 'Decile')
```

Income levels generally experienced healthy growth from year 2004 onwards.
The bottom income earners actually suffered stagnant income growth all the way till year 2009
and have significantly lagged behind the other deciles as a result.

It can also be observed that the extent income growth of each decile tend to follow decile levels,
with higher deciles experiencing better growth than lower deciles.

The total and compounded annual growth rates are presented in the table below.

```{r decile.growth, results = 'asis'}
income.growth <- income.decile %>%
                    group_by(decile) %>%
                    summarise(growth.total = (last(index.2000) - 1) * 100,
                              growth.annual = (last(index.2000) ^ (1/13) - 1) * 100) %>%
                    ungroup()
kable(income.growth, digits = 2, caption = 'Table of Growth Rates between Years 2000 and 2013 by Deciles')
```

Next let's look at the distribution of income relative to the total income to investigate 
how it has changed over the years.

```{r decile.proportion}
# remove average values and compute proportion numbers
income.decile <- income.decile %>%
                    filter(decile %in% 1:10) %>%
                    mutate(decile = as.numeric(levels(decile)[decile]),
                           population.cumprop = decile / 10)
income.decile <- income.decile %>%
                    group_by(year) %>%
                    arrange(decile) %>%
                    mutate(income.prop = income / sum(income),
                           income.cumprop = cumsum(income.prop)) %>%
                    ungroup()
```

### Line Graph of Income Proportion by Deciles

First, a line graph for simple comparisons

```{r decile.prop.line, results = 'asis'}
income.decile %>%
    ggvis(~factor(year), ~income.prop, stroke = ~factor(decile)) %>%
    group_by(decile) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Income Proportion', title_offset = 50) %>%
    add_legend('stroke', title = 'Decile')
```

Despite the unequal distribution of income, it can be observed from this graph that the distribution
has been quite consistent through the years.
Except for the slight increase in income proportion of the top decile earners,
the income proportion is quite flat for each decile from year 2000 to year 2013.

### Stacked Area Graph of Income Proportion by Deciles

The unequal distribution can be more easily observed in a stacked area graph.

```{r decile.stackedarea, results = 'asis'}
income.decile %>%
    group_by(year) %>%
    mutate(to = income.cumprop, from = c(0, to[-n()])) %>%
    ggvis(x = ~factor(year), fill = ~factor(decile), opacity := 0.8) %>%
    group_by(decile) %>%
    layer_ribbons(y = ~from, y2 = ~to) %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Cumulative Income Proportion') %>%
    add_legend('fill', title = 'Decile')
```

It can be observed that the top decile earners make more than 
the total income of the bottom 5 deciles earners.
In other words, the top 10% of earners made more income than the bottom 50% of income earners. WOW!
Similar comparisons can be made with top 2 vs bottom 6 deciles and top 3 vs bottom 7 deciles.
The top 30% of earners are responsible for more than half the total income of Singapore earners.

The cumulative proportion of income in Singapore from the bottom decile is show in the table below.

```{r decile.cumprop, results = 'asis'}
income.decile %>%
    filter(year == 2013) %>%
    select(decile, income.cumprop) %>%
    mutate(income.cumprop = income.cumprop * 100) %>%
    kable(digits = 2, caption = 'Table of Cumulative Proportion of Income by Deciles in 2013')
```

### Lorenz Curve

The unequal distribution can also be shown in a [Lorenz curve](https://en.wikipedia.org/wiki/Lorenz_curve), 
which plots the cumulative income proportion against cumulative population proportion.
The straight line joining zero and one shows perfect equality of income distribution.

```{r decile.lorenz, results = 'asis'}
income.decile %>%
    select(year, population.cumprop, income.cumprop) %>%
    rbind(data.frame(year = 2000:2013, population.cumprop = 0, income.cumprop = 0)) %>%
    rbind(data.frame(year = NA, population.cumprop = c(0, 1), income.cumprop = c(0, 1))) %>%
    ggvis(~population.cumprop, ~income.cumprop, stroke = ~factor(year)) %>%
    group_by(year) %>%
    layer_lines() %>%
    add_axis('x', title = 'Cumulative Population Proportion') %>%
    add_axis('y', title = 'Cumulative Income Proportion', title_offset = 40) %>%
    add_legend('stroke', title = 'Year', values = 2000:2013)
```

Consistent with the observation with the stacked area graph, the distribution of income
is fairly contant as shown by the well overlayed curves of 14 years of data.

### Ratio to Average Income of Bottom 10% by Deciles

Another way to show unequal income distrubution in a country is the ratio of 
the average income of the top 10 percentile to that of the bottom 10 percentile.
I have done so for all the deciles to the bottom decile as follows.

```{r decile.multiple, results = 'asis'}
# compute ratio of average income to bottom decile
income.decile <- income.decile %>%
                    group_by(year) %>%
                    arrange(decile) %>%
                    mutate(bottom.multiple = income / first(income)) %>%
                    ungroup()
income.decile %>%
    select(year, decile, bottom.multiple) %>%
    ggvis(~factor(year), ~bottom.multiple, stroke = ~factor(decile)) %>%
    group_by(decile) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Multiple') %>%
    add_legend('stroke', title = 'Decile')
```

It can be observed that the top decile earned on average
`r with(income.decile, round(bottom.multiple[year == 2013 & decile == 10], 2))` times
the average income of the bottom decile in 2013.
Even the average income 2nd highest decile was 
`r with(income.decile, round(bottom.multiple[year == 2013 & decile == 9], 2))` times
the bottom decile in 2013.
While the ratios to the bottom decile had increase from year 2000, 
the consolation is that they has been relatively flat from around 2008 and 2009 onwards.
This meant that the bottom decile has been able to grow their income at the same proportion as the other deciles.

### Ratio to Average Income of Bottom 20% by Quintiles

It is usually the case that the ratio of income is made between the top 20% against the bottom 20%,
which will be what I attempt next.

```{r quintile.multiple, results = 'asis'}
# create quintile groups and compute mean income
income.quintile <- income.decile %>%
                        mutate(quintile = (decile + 1) %/% 2) %>%
                        group_by(year, quintile) %>%
                        summarise(income = mean(income)) %>%
                        ungroup()
# compute ratio of average income to bottom quintile
income.quintile <- income.quintile %>%
    group_by(year) %>%
    arrange(quintile) %>%
    mutate(bottom.multiple = income / first(income)) %>%
    ungroup()
income.quintile %>%
    select(year, quintile, bottom.multiple) %>%
    ggvis(~factor(year), ~bottom.multiple, stroke = ~factor(quintile)) %>%
    group_by(quintile) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Multiple') %>%
    add_legend('stroke', title = 'Quintile')
```

The general trend is the same, with the top 20% earning 
`r with(income.quintile, round(bottom.multiple[year == 2013 & quintile == 5], 2))` times
the average income of the bottom 20% of earners in 2013.

## Distribution of Households by Monthly Income

Finally, I will explore the distribution of households by monttly income in Singapore.
The raw data is shown in the table as follows.

```{r distribution.data, results = 'asis'}
income.distribution <- read.table('households-by-monthly-income.txt', header = T, check.names = F)
kable(income.distribution, digits = 2, caption = 'Table of Household Proportion by Monthly Income')
```

Some data transformation is necessary as the umemployed proportions are not the interest of this exercise.
I would exclude the percentages and scale up the rest of the percentages accordingly.

```{r distribution.transform}
# convert to long format
income.distribution <- income.distribution %>%
                            gather(income, percentage, -year)
# exclude unemployed and rescale proportion
income.distribution <- income.distribution %>%
                            filter(income != 'unemployed') %>%
                            droplevels() %>%
                            group_by(year) %>%
                            mutate(proportion = percentage / sum(percentage)) %>%
                            ungroup()
```

### Line Graph of Household Proportion by Income Grooup

Due to the large number of groups which also follow different trends, I have elected to split this into
2 graphs.

```{r distribution.line1, results = 'asis'}
# for monthly income groups less than 7000
income.distribution %>%
    filter(income %in% levels(income)[1:7]) %>%
    droplevels() %>%
    ggvis(~factor(year), ~proportion, stroke = ~income) %>%
    group_by(income) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Population Proportion', title_offset = 40) %>%
    add_legend('stroke', title = 'Income Group')
```

This graph shows the proportion of households with monthly income less than 7000.
The proportions in these income groups were generally decreasing, which is expected given
the income growth experienced. 
The proportion of households earning below 1000 fell below the year 2000 levels only
after year 2009, however, it started at a very low base of 
`r with(income.distribution, round(proportion[income == levels(income)[1] & year == 2000] * 100, digits = 2))`%

```{r distribution.line2, results = 'asis'}
# for monthly income groups more than 7000
income.distribution %>%
    filter(income %in% levels(income)[8:18]) %>%
    droplevels() %>%
    ggvis(~factor(year), ~proportion, stroke = ~income) %>%
    group_by(income) %>%
    layer_lines() %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Population Proportion', title_offset = 40) %>%
    add_legend('stroke', title = 'Income Group')
```

On the other hand, the proportions for the groups with monthly income above 7000 were generally increasing.
The most exaggerated increase belonged to the group for the highest income group: 
those making more than 20,000 per month.

### Stacked Area Graph of Household Proportion by Monthly Income

I think it may be more intersting to observe this distribution of household by a stacked area graph.

```{r distribution.stackedarea, results = 'asis'}
income.distribution %>%
    group_by(year) %>%
    mutate(to = cumsum(proportion), from = c(0, to[-n()])) %>%
    ggvis(x = ~factor(year), fill = ~factor(income), opacity := 0.8) %>%
    group_by(income) %>%
    layer_ribbons(y = ~from, y2 = ~to) %>%
    add_axis('x', title = 'Year') %>%
    add_axis('y', title = 'Cumulative Population Proportion') %>%
    add_legend('fill', title = 'Income Group')
```

Not the most informative graph as the accumlated effects of the lower income groups makes it difficult
to make inference about the trends of the higher income groups.
A simple observation would be that about 50% of the households made less than 7000 monthly in 2013.
This is consistent with the median income level being 7000 monthly.
The income groups that formed the largest proportion of households are those between 1000 and 7000
and the higher earners above 20,000.

The cumulative proportion of households by income group in 2013 is shown in the table below.

```{r distribution.cumprop, results = 'asis'}
income.distribution %>%
    filter(year == 2013) %>%
    mutate(cumprop = cumsum(proportion) * 100) %>%
    select(income, cumprop) %>%
    kable(digits = 2, caption = 'Table of Cumulative Proportion of Households by Income Group in 2013')
```

### Household Distribution by Monthly Income

Finally, let's look at the change in household distribution by monthly income over the years.

```{r distribution.year, results = 'asis'}
income.distribution %>%
    ggvis(~income, ~proportion, stroke = ~factor(year)) %>%
    group_by(year) %>%
    layer_lines() %>%
    add_axis('x', title = 'Income Group', title_offset = 50,
             properties = axis_props(labels = list(angle = 20, align = 'left'))) %>%
    add_axis('y', title = 'Population Proportion', title_offset = 40) %>%
    add_legend('stroke', title = 'Year')
```

As evident in the previous graphs, the proportion of households with income less than 6000 have been decreasing,
with the proportions moving to income groups more than 7000.
The group earning between 2000 and 3000 have traditionally formed the largest proportion of household in Singapore.
However, the proportion of household earning more than 20,000 formed the largest proportion in the latest
data in year 2013.

## Conclusion

I hope this visualisation exercise of household income data of Singapore has given some insights
into the distribution and trends of income in Singapore.
It has definitely been a good practice for me in using the data wrangling and visualisation packages
by [Hadley Wickham](http://had.co.nz/) (i.e. `dplyr`, `tidyr`, `ggvis`) 
and also writing reports in R Markdown.